Visualizing probability distributions across bivariate cyclic temporal granularities

### 2020-03-15 main.Rmd
revisions till cyclic time granularities

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities."}
knitr::include_graphics("Figs/linear-arr.png")
```
In addition to the integer index, each granule could be associated with a textual representation which is called the label.<!--The label can be mapped to index for input and the index can later be mapped to this label for output.--> When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.-->

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] framework of organizing temporal data. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->

@@

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap="Circular granularity day-of-week"}
knitr::include_graphics("Figs/circular-dow.png")
```

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
minute-of-hour: & $C_1$ & $= L_1(z \mod 60)$ & $P_1 =$&$60$ \\
minute-of-day: & $C_2$ & $= L_2(z \mod 60*24)$ & $P_2=$&$`r 60*24`$\\
hour-of-day: & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
hour-of-week: & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
day-of-week: & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$ & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with time index in minutes}
\label{tab:definitions}
\end{table}

@@

day-of-month: & $Q_1$ && $P_1 = 31, 30, 29, 28$\\
hour-of-month: & $Q_2$ && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
day-of-year: & $Q_3$ && $P_3 = 366, 365$\\
week-of-month: & $Q_4$ && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative quasi-circular granularities with potential period lengths}
\label{tab:quasi}
\end{table}

@@

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap="Quasi-circular granularities $Q(B, G')$ and $Q(B, H')$ relating bottom granularity $B$ and linear granularities $G'$ and $H'$ respectively."}
knitr::include_graphics("Figs/quasi-circular-example.png")
```

@@

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$. <!--that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity.--> Further assume that $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

\begin{definition}[Aperiodic cyclic granularity]\label{aperiodic}
An aperiodic cyclic granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}(j)$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="Aperiodic cyclic granularity $A(B, M1)$ relating aperiodic linear granularity $M1$ and bottom granularity $B$"}
knitr::include_graphics("Figs/aperiodic-example.png")
```

>>>>

```{r linear-time, echo=FALSE, out.width = "100%", fig.cap="The time domain distributed as linear granularities hours, days, weeks and months. These are ordered with ordering guided by integers and hence is unidirectional and non-repeating."}
knitr::include_graphics("Figs/linear-arr.png")
```
In addition to the integer index, each granule could be associated with a textual representation which is called the label.<!--The label can be mapped to index for input and the index can later be mapped to this label for output.--> When the label set is exactly the integers, the granularities are called "full-integer" labeled granularity, such as in Definition \ref{def:linear}.<!-- Some calendar operations will only work on full-integer labeled granularities, while others can be implemented using more flexible labelings.-->

@@

We propose a formalism of cyclic time granularities through the tsibble [@wang2019tsibble] data structure. A tsibble consists of an index, key and measured variables. An index is a variable with inherent ordering from past to present and a key is a set of variables that define observational units over time. <!--In a tsibble, each observation (row) is uniquely identified by index and key.--> A time domain, as defined by @Bettini1998-ed, is essentially a mapping of the index set to the time index of a tsibble. A linear granularity is a mapping of the index set to subsets of the time domain. For example, if the time index of a tsibble is days, then a linear granularity might be weeks, months or years. A bottom granularity is represented by the index of the tsibble.<!--Consistent with the assumption of @Bettini2000-vy, all linear granularities can be generated from the bottom granularity by calendar algebraic operations.-->

@@

```{r circular-dow,echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear granularities Day and Week and circular granularity Day-of-Week with $P=7$ is constructed using the two linear granularities. Day-of-Week has a repeating pattern unlike Day and Week and has an ordering only within a period represented by hashed arrows."}
knitr::include_graphics("Figs/circular-dow-arrow.png")
```

@@

\begin{table}[ht]
\begin{center}
\begin{tabular}{lr@{~}lr@{~}r}
\toprule
minute-of-hour: & $C_1$ & $= L_1(z \mod 60)$ & $P_1 =$&$60$ \\
minute-of-day: & $C_2$ & $= L_2(z \mod 60*24)$ & $P_2=$&$`r 60*24`$\\
hour-of-day: & $C_3$ & $= L_3(\lfloor z/60\rfloor\mod 24)$  & $P_3 =$&$24$ \\
hour-of-week: & $C_4$ & $= L_4(\lfloor z/60\rfloor\mod 24*7)$  & $P_4=$&$`r 24*7`$\\
day-of-week: & $C_5$ & $= L_5(\lfloor z/24*60\rfloor \mod 7)$ & $P_5=$&$7$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative circular granularities with bottom granularity "minutes". These circular granularities relate two linear granularities one of which groups periodically into the other with regular mapping.}
\label{tab:definitions}
\end{table}

@@

day-of-month: & $Q_1$ && $P_1 = 31, 30, 29, 28$\\
hour-of-month: & $Q_2$ && $P_2 = 24*31, 24*30, 24*29, 24*28$\\
day-of-year: & $Q_3$ && $P_3 = 366, 365$\\
week-of-month: & $Q_4$ && $P_4 = 5, 4$\\
\bottomrule
\end{tabular}
\end{center}
\caption{Illustrative quasi-circular granularities with potential period lengths. These quasi-circular granularities relate two linear granularities one of which groups periodically into the other with irregular mapping.
Illustrative quasi-circular granularities }
\label{tab:quasi}
\end{table}

@@

```{r quasi-circular-example, echo = FALSE, out.width = "100%",fig.cap="The time domain distributed as linear granularities $B$, $G'$, $H'$ and quasi-circular granularities $Q(B, G')$ and $Q(B, H')$. Both $Q(B, G')$ and $Q(B, H')$ have repeating patterns unlike $G'$ or $H'$ and the patterns are guided by the distribution of $B$ within $G'$ and $H'$ within a period shown by the hashed arrow."}
knitr::include_graphics("Figs/quasi-circular-ex.png")
```

@@

Aperiodic cyclic granularities are defined using aperiodic linear granularities. Consider n aperiodic linear granularities $M_i$  $\forall \{i\in{1, 2, \dots, n\}}$ such that \ref{def:periodical} does not hold true for them with respect to $B$ (the bottom granularity). <!--that is, none of them can be expressed as periodic or quasi-periodic with respect to the bottom granularity.--> However, $B \trianglelefteq M_i$ $\forall \{i\in{1, 2, \dots, n\}}$. Then according to Definition \ref{def:groupsinto}, for each index $j$ there exists a (possibly infinite) subset $T_{\{i_j\}}$ of the integers such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$ and suppose $M = \bigcup_{i=1}^{n}M_{i}$ is formed by collecting the granules of $\{M_1, M_2, \dots, M_n\}$. Here, index ${\{i_j\}}$ stands for the $j^{\text{th}}$ granule of the $i^{\text{th}}$ linear aperiodic granularity.

\begin{definition}[Aperiodic cyclic granularity]\label{aperiodic}
An aperiodic cyclic granularity $A_{B, M}$ relates aperiodic linear granularity $M$ and bottom granularity $B$, if
\begin{equation}\label{eq:aperiodic}
\begin{split}
A_{B, M}(z) & = L(i), \quad z \in T_{i_j} \quad and \quad \forall i \in {1, 2, \dots, n}\\
     & = L(0), \quad otherwise
\end{split}
\end{equation}
where,
$z \in \mathbb{Z}_{\geq 0}$ denotes the index set,
$T_{i_j}$ are the sets of indices of $B$ describing aperiodic linear granularities $M_{i}$ such that $M_{i}(j) = \bigcup_{z \in T_{i_j}}B(z)$, $M = \bigcup_{i=1}^{n}M_{i}$.
\end{definition}

@@

```{r aperiodic-example, echo=FALSE,out.width = "100%",fig.cap="The time domain distributed as linear aperiodic granularities $B$, $M1$, $M2$, $M$ and aperiodic cyclic granularity $A(B, M)$. $A(B, M)$ have repeating patterns with three categories, two of which correspond to the aperiodic linear granularities $M1$ and $M2$ and other one correponding to no aperiodic granularity."}
knitr::include_graphics("Figs/aperiodic-ex.png")
```
